---
title: "predmachlearn-006-project"
author: "Knut Behrends"
date: "22.10.2014"
output: html_document
---

```{r echo=FALSE}
if(system("uname -n") == "well"){
        setwd("~/Documents/coursera/datascience/08-machinelearning/predmachlearn-006-project")
}

```
## Predicting the quality of physical exercising.


Analysis from sensor data acquired during exercising with barbells.

Please read the file [Background](background.html) to get familiar with the subject matter.

### Read In Data Files, Load R Packages
```{r init}
training <- read.csv("../data/pml-training.csv")
testing <- read.csv("../data/pml-testing.csv")


```

```{r return="hide", message=FALSE}
library(Hmisc)
#contents(training)
#descr <- describe(training)
#str(descr)



```
There is not much data cleaning necessery. All columns have nice all-lowercase, self-describing names, and few datatypes : `r unique(sapply(training, class))`).
There are `r length(names(training))` variables: 

```{r }
nas <- colSums(is.na(training));table(nas)

```
There are `r length(nas[nas == 0 ])` columns which have no NA values, and `r length(nas[nas !=0 ]) ` columns all of which have the same number of `r sprintf("%i", max(nas[nas !=0 ]))` NAs.